Foregrounds in Fourier Delay Space for Wide--Field Redshifted 21~cm Observatories

### 2015-02-08 ms.tex
Make changes suggested by Judd
from eb548c5:
============

In ms.tex,

(*) Change title to mention "Fourier delay space" and "Wide-field"
observatories
(*) Change abstract slightly to mention "Fourier delay space"
(*) Change introduction slightly
(*) Change Delay Spectrum section slightly
(*) Change Wide-field effects section slightly
(*) Re-number subsections in Simulations section and change slightly
(*) Trim "Analysis of Foreground Signatures" section by a few paragraphs
and change "MWA Data Modeling" section slightly
(*) Change title and wording of "Baseline based foreground mitigation"
se

%% \title{Detecting Epoch of Reionization in redshifted 21~cm: contamination in EoR window}
%% \title{Limits on the detection of the Epoch of Reionization from MWA observations of the redshifted 21~cm line}
\title{Foregrounds in Fourier Space for Redshifted 21~cm Observatories}

@@

Detection of 21~cm emission of H{\sc i} from the epoch of reionization, at redshifts $z>6$, is limited primarily by foreground emission. We investigate the signatures of wide--field measurements and an all--sky foreground model using the delay spectrum technique that maps the measurements to foreground object locations through signal delays between antenna pairs. We demonstrate interferometric measurements are inherently sensitive to all scales, including the largest angular scales, owing to the nature of wide--field measurements. These wide--field effects are generic to all observations but antenna shapes impact their amplitudes substantially. A dish--shaped antenna yields the most desirable features from a foreground contamination viewpoint, relative to a dipole or a phased array. Comparing data from recent Murchison Widefield Array observations, we demonstrate that the foreground signatures that have the largest impact on the H{\sc i} signal arise from power received far away from the primary field of view. We identify diffuse emission near the horizon as a significant contributing factor, even on wide antenna spacings. For signals entering through the primary field of view, compact emission dominates the foreground contamination. These two mechanisms imprint a characteristic {\it pitchfork} signature on the ``foreground wedge'' in Fourier space. Based on these results, we propose that selective down--weighting of data based on antenna spacing and time can mitigate foreground contamination substantially by a factor $\sim 100$ with negligible loss of sensitivity.

@@

At the end of the recombination epoch, the Universe was completely neutral. This period, referred to as the {\it Dark Ages} in the Universe's history, is characterized by the localized accumulation of matter under the influence of gravity. And it ended with the formation of the first stars and galaxies which started emitting ultra--violet and X--ray radiation, thereby reionizing the neutral medium in their surroundings. This commenced the epoch of reionization (EoR) --- a period of non--linear growth of matter density perturbations and astrophysical evolution. Studying the EoR holds the key to understanding this evolution.

Observations of redshifted 21~cm radiation generated by the spin flip transition of H{\sc i} has been identified as a direct probe of the EoR \citep{sun72,sco90,mad97,toz00,ili02}. Detecting this signal has recently emerged as a very promising experiment to fill the gaps in our understanding of the Universe's history.

@@

Considerable effort has been and continues to be made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum and formulating robust estimators of the true power spectrum \citep{bow09,liu09,dat10,liu11,mor12,tro12,pob13,thy13,dil14,liu14a,liu14b}. {\bf A model that provides a generic explanation for the observed foreground power spectrum has emerged, whereby the wide--field (and chromatic) response of the instrument causes the power in smooth spectrum foregrounds to occupy higher $k$--modes into the so-called `wedge'}. % A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide--field (and chromatic) response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. 
The conservative foreground strategy, referred to as {\it avoidance}, that has developed alongside this work is to discard $k$--modes which could be contaminated \citep[for e.g.,][]{par14}. The more aggressive alternative is to subtract a sky model and regain access to modes discarded by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we focus primarily on extending the {\it avoidance} strategy by identifying foreground components at greatest risk to ``leak'' from foreground modes to EoR modes and proposing a scheme for down--weighting these components.

Foregrounds with intrinsic deviations from spectral smoothness, instruments with high chromaticity, polarization leakage, calibration errors, or approximations in power spectrum analyses can contaminate the true EoR H{\sc i} power spectrum. Here we use existing catalogs and a high fidelity instrumental model to capture both foreground and instrumental chromaticity. To decouple these effects from possible analysis effects, such as those pointed out by \citet{haz13}, we compute power spectra using a {\it \textbf{Per-Baseline}} approach of \citet{par12b}. This approximates the power spectrum as the inverse Fourier transform of the spectra generated by the instrument's correlator.

{\bf We investigate signatures generic to all wide--field measurements of EoR power spectrum and explore antenna shapes that are most desirable from the viewpoint of foreground contamination for future EoR experiments}. After matching observations from the Murchison Widefield Array with instrumental simulations of the entire sky, we examine the {\bf shape of the delay spectrum} of each baseline. We report two important findings: the foregrounds that most severely {\bf obscure} the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon; and, diffuse Galactic emission plays a significant role hitherto unpredicted. We quantify these by separating the simulations into components based on type of foreground emission. We then arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to down--weight adversely contaminated baselines.

@@

Interferometer array data known as {\it visibilities}, $V_{bf}\equiv V_{bf}(\boldsymbol{b},f)$, represent correlations between time--series of electric fields measured by different antenna pairs with separation vectors $\boldsymbol{b}$ and then Fourier transformed along the time axis to obtain a spectrum along the frequency ($f$) axis. If $I(\hat{\boldsymbol{s}},f)$ and $A(\hat{\boldsymbol{s}},f)$ are the sky brightness and antenna's directional power pattern, respectively, at different frequencies as a function of direction on the sky denoted by the unit vector ($\hat{\boldsymbol{s}}$), and $W_f\equiv W_f(f)$ denotes instrumental bandpass weights, then $V_{bf}$ can be written as (with a slight adaptation from \citet{van34}, \citet{zer38}, and \citet{tho01}):
\begin{align}\label{eqn:obsvis}
  V_{bf} &= \iint\limits_\textrm{sky} A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)\,W_f\,e^{-i2\pi f\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}}\,\dif\Omega,
\end{align}
where, $c$ is the speed of light, and $\dif\Omega$ is the solid angle element to which $\hat{\boldsymbol{s}}$ is the unit normal vector. We wish to {\bf note} that this equation is valid {\bf in general, including wide--field measurements}.

@@

Equation~\ref{eqn:obsvis} can be {\bf equivalently expressed} as:
\begin{align}\label{eqn:vis}
  V_{uf} &\equiv V_{uf}(\boldsymbol{u},f) \nonumber\\
  &= \iint\limits_\textrm{sky} A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)\,W_f\,e^{-i2\pi\boldsymbol{u}\cdot\hat{\boldsymbol{s}}}\,\dif\Omega,
\end{align}
where, $\hat{\boldsymbol{s}}$ is measured with reference to a location on the sky referred to as the {\it phase center}, and $\boldsymbol{u}\equiv (u,v,w)$ denotes the spatial frequency vector. $w$ is aligned parallel to the direction of the phase center, while $u$ and $v$ lie on the transverse plane perpendicular to it. For measurements that are constrained to lie on this plane, we can choose $w=0$ without loss of generality and $\boldsymbol{u}$ effectively reduces to $\boldsymbol{u}\equiv (u,v)$, a two--dimensional vector. Then, $\boldsymbol{u}$ is directly related to the transverse spatial wavenumber mode as: 
\begin{align}\label{eqn:k-perp}
  \boldsymbol{k}_\perp &\equiv \frac{2\pi\boldsymbol{u}}{D}, 
\end{align}
where, $D\equiv D(z)$ is the transverse comoving distance at redshift $z$. % Hence, $V_{uf}=V_{uf}(\boldsymbol{b}/\lambda,f)=V_{bf}(\lambda\boldsymbol{u},f)$. It is important to note that for a given antenna pair, $\boldsymbol{b}$ is independent of frequency while $\boldsymbol{u}$ is not.

@@

% Radio emission from foreground objects, due to their spatial and spectral (chromatic) structure, contaminate the received signal \citep{bow09,dat10,mor12,tro12,thy13,liu14a,liu14b}. 
Foregrounds can be described in both the delay spectrum and Fourier frameworks. For our study, which primarily concerns with foreground characterization, we find the former to be simple and yet extremely useful, while maintaining a close correspondence with the latter despite subtle differences. In order to express a quantity derived from $V_{u\eta}$ that has units identical to that of the cosmological H{\sc i} power spectrum, we define delay power spectrum:
\begin{align}\label{eqn:fake-power-spectrum}
  P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel) &\equiv |V_{b\tau}|^2\left(\frac{A_\textrm{e}}{\lambda^2\Delta B}\right)\left(\frac{D^2\Delta D}{\Delta B}\right)\left(\frac{\lambda^2}{2k_\textrm{B}}\right)^2,
\end{align}
where, $\boldsymbol{u}$ and $\eta$ in equations~\ref{eqn:k-perp} and \ref{eqn:k-prll} have been correspondingly replaced with $\boldsymbol{b}/\lambda$ and $\tau$ respectively.

In summary, the delay spectrum, $V_{b\tau}$, is obtained from {\it visibilities} which are the basic data blocks measured by each antenna pair, using equations~\ref{eqn:obsvis} and \ref{eqn:delay-transform}. $V_{b\tau}$ captures all the effects of EoR H{\sc i} signal corruption caused by foregrounds and the instrument. At the same time, it is closely related to the sought power spectrum $P(\boldsymbol{k}_\perp,k_\parallel)$ containing critical information about spatial scales. % In this paper, we often use the quantity $P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel)$, and sometimes $V_{b\tau}$.

@@

{\bf We give a brief overview of some parameters of Fourier space which is generic to all experiments that use a similar approach}. Figure~\ref{fig:fourier-space} illustrates the Fourier space in which the delay (and power) spectra of H{\sc i} from the EoR are presented in this paper. $|\boldsymbol{b}|$ and $k_\perp$, denoting spatial scales in the transverse direction {\bf (tangent plane to the celestial sphere)}, form the $x$--axis. $\tau\approx\eta\propto k_\parallel$, denoting spatial scales along line--of--sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge} \citep{dat10}, whose boundaries are determined by the antenna spacings and the light travel times across them. These boundaries, called horizon delay limits \citep{ved12,par12b}, are shown by solid lines. The spectral transfer function of the instrument convolves the {\it foreground wedge} and stretches it further (unshaded narrow strips bounded by solid and dashed lines) along $\tau$--axis \citep{par12b,thy13}. The width of this narrow strip is inversely proportional to the operating bandwidth. The region of Fourier space excluding the {\it foreground wedge} and the narrow strips is the so--called {\it EoR window}, shown in light and medium shades. {\bf In the context of EoR studies in Fourier space, the H{\sc i} power spectrum from the EoR is expected to decrease rapidly with $|\boldsymbol{k}|$. Hence, the brightest EoR signal will be observed on the shortest baselines and smallest delays. Thus the region of interest for EoR studies is just beyond the horizon delay limits (dashed lines) on short baselines, marked as the region of {\it maximal EoR sensitivity}}.

{\bf In the specific case of the MWA which has a passband constructed using coarse channels, there are period grating responses resulting in repetitions of the {\it foreground wedge} at multiples of 0.78~$\mu$s. Thus with reference to the MWA, the {\it EoR window} is assumed to lie outside the dashed lines but inside the first grating response ($|\tau| \lesssim 0.78\,\mu$s, dot--dashed lines) and is shown in medium shade. The MWA EoR collaboration actively uses this as a test region in Fourier space for diagnostics on data quality}.

@@

We obtain the delay spectrum of visibilities by taking the delay transform of each baseline's spectrum (equation~\ref{eqn:delay-transform}) choosing $W'_f$ to be a {\it Blackman--Harris} window function. The spectrum is multiplied by the additional weights, $W_f$, consisting of instrumental passband and channels possibly flagged for interference. In delay--space, these weights translate into a convolution by a point spread function (PSF). We deconvolve this PSF using a one dimensional CLEAN algorithm \citep{tay99} as described for the delay axis \citep{par09,par12b} to obtain the final delay spectra. The CLEAN procedure iteratively finds and subtracts peak values convolved by the Fourier transform of the weights. We limit the selection of peaks to modes inside the horizon delay limit, corresponding to smooth spectrum objects in the visible sky hemisphere.

@@

When the synthesized field is small, where $A(\hat{\boldsymbol{s}},f)$ or $I(\hat{\boldsymbol{s}},f)$ is negligible for $|l| \ll 1$ and $|m| \ll 1$, equation~\ref{eqn:obsvis-lmn} reduces to a simple two--dimensional Fourier transform \citep{tay99,tho01} between the apparent sky brightness and measured visibilities. It is in this context that radio interferometers are understood to be sensitive only to fluctuations and not to a uniform sky brightness distribution.

In a wide--field measurement, neither $A(\hat{\boldsymbol{s}},f)$ nor $I(\hat{\boldsymbol{s}},f)$, in general, is negligible anywhere in the visible hemisphere. The solid angle per pixel on the sky in direction cosine coordinates changes significantly with direction (equation~\ref{eqn:solid-angle}), increasing steeply towards the horizon. Hence, the approximations in the narrow--field scenario do not apply. For example, even if $A(\hat{\boldsymbol{s}},f)$ and $I(\hat{\boldsymbol{s}},f)$ are held constant across the visible hemisphere, the integrand in equation~\ref{eqn:obsvis-lmn} is still dependent on direction. Therefore, in a significant departure from a narrow--field measurement, the wide--field visibility from a uniform brightness distribution on a non--zero antenna spacing is not zero.

@@

We interpret this is due to equal--sized delay bins subtending larger solid angles near the horizon thereby containing larger integrated emission. An alternate, but equivalent, interpretation is that baseline vectors (including those with largest lengths) are foreshortened towards the horizon along their orientation. Thus, they become sensitive to larger angular scales that match the inverse of their foreshortened baseline lengths along these directions.

\citet{thy13} have found evidence of this feature in their statistical models. In line with our reasoning, they attribute it to a steep rise in solid angles subtended by delay bins near the horizon limits. \citet{pob13} also find a similar ``edge brightening'' feature which they attribute to emission from the Galactic plane near the horizon. From their discussion, it is unclear what fraction of power in that feature arises from such wide--field effects.

We conclude these are generic to all instruments making wide--field measurements. The nature of the specific instrument used for observing will control the amplitude of these effects, which we explore below. % Typically, the foreground sky consists of diffuse emission and point-like sources, and the antenna is more sensitive to power received from primary field of view than from the rest of the sky.

@@

\subsubsection{Diffuse Foreground Model}\label{sec:DSM}

For the diffuse component, we use an all--sky radio foreground model \citep{deo08} to estimate the emission at 185~MHz. At this frequency, since this map is predominantly based on the 408~MHz map of \citet{has82} which has an angular resolution of 0.85\arcdeg, we smoothed the 185~MHz map to the same resolution. However, to avoid any artifacts from sampling this map, we sample it at $\approx 27$\arcmin~intervals. We model the diffuse foreground spectra with a spectral index at each pixel in the map, estimated from model maps at 170~MHz and 200~MHz.

\subsubsection{Compact Foreground Model}\label{sec:CSM}

The model described above is primarily a model of the diffuse foreground sky. While it contains faint compact emission blended in with the diffuse emission, {\bf bright point sources have been removed \citep{deo08}}. In order to supplement it with missing bright compact emission, we use classical radio source confusion estimates to determine the nominal flux density threshold and include point sources brighter than this threshold. It may be noted that slightly different criteria are in common use in radio astronomy to estimate radio source confusion \citep[see Appendix of][and references therein]{thy13}. For an angular resolution of 0.85\arcdeg, using a conservative `$S_\textrm{c}=5\sigma_\textrm{c}$' criterion, we determine the flux density threshold to be $\approx 10$~Jy. Other liberal criteria that yield a lower threshold carry a greater risk of double--counting point sources which might be already blended in with the diffuse sky model.

@@

Based on these criteria, we select 100 objects from the SUMSS catalog and 250 objects from the NVSS catalog. Together with the diffuse foreground model, we obtain an all--sky foreground model consisting of both compact and diffuse emission. Figure~\ref{fig:sky-model-generic} shows the diffuse (top) and compact (bottom) foreground emission model used in our study. In this snapshot at 0.09~hours LST, the Galactic center in the diffuse model appears to have just set in the West.

@@

The instrument model consists primarily of the antenna power pattern, $A(\hat{\boldsymbol{s}},f)$ (see equation~\ref{eqn:obsvis}). It is determined by the shape of its aperture. Through the use of a few examples, we examine the role the geometrical shape of the aperture plays in shaping the characteristics of delay power spectrum. We consider the following antenna elements placed at the MWA tile locations:
\begin{itemize}
\item {\it Dipole}: An East--West dipole of length 0.74~m at a height 0.3~m above a ground plane. $A_\textrm{e}=(\lambda/2)^2$.
\item {\it Phased Array}: A 4$\times$4 array of isotropic radiators with a grid spacing of 1.1~m at a height 0.3~m above the ground plane placed in an arrangement similar to that in an MWA tile. $A_\textrm{e}=16\,(\lambda/2)^2$. 
\item {\it Dish}: Diameter of 14~m similar to that proposed for HERA, with $A_\textrm{e}\approx 154$~m$^2$. The power pattern is simulated using an {\it Airy} pattern where its sensitivity beyond the horizon is forced to zero.
\end{itemize}

The power patterns of these antenna geometries at 185~MHz for the {\it zenith} pointing are shown in Figure~\ref{fig:power-patterns}. It must be noted that the presence of a ground plane below the dipole and the phased array introduces a smoother suppression of sensitivity near the horizon, whereas this transition is abrupt in the case of the dish.

@@

The delay power spectra without thermal noise component for these antenna shapes are shown in Figure~\ref{fig:antennas-delay-spectra}. The occupancy of the power patterns on the sky is clearly correlated with that in the delay spectra. The following visual correlations are also apparent --- the strength of the primary lobe centered on the pointing center is correlated with the delay power spectrum centered on $\tau=0$; and, the overall rate of decrease in the power sensitivity away from the pointing center is correlated with the rate of drop in power away from $\tau=0$.

>>>>

%% \title{Detecting Epoch of Reionization in redshifted 21~cm: contamination in EoR window}
%% \title{Limits on the detection of the Epoch of Reionization from MWA observations of the redshifted 21~cm line}
\title{Foregrounds in Fourier Delay Space for Wide--Field Redshifted 21~cm Observatories}

@@

Detection of 21~cm emission of H{\sc i} from the epoch of reionization, at redshifts $z>6$, is limited primarily by foreground emission. We investigate the signatures of wide--field measurements and an all--sky foreground model using the delay spectrum technique that maps the measurements to foreground object locations through signal delays between antenna pairs. We demonstrate interferometric measurements are inherently sensitive to all scales, including the largest angular scales, owing to the nature of wide--field measurements. These wide--field effects are generic to all observations but antenna shapes impact their amplitudes substantially. A dish--shaped antenna yields the most desirable features from a foreground contamination viewpoint, relative to a dipole or a phased array. Comparing data from recent Murchison Widefield Array observations, we demonstrate that the foreground signatures that have the largest impact on the H{\sc i} signal arise from power received far away from the primary field of view. We identify diffuse emission near the horizon as a significant contributing factor, even on wide antenna spacings. For signals entering through the primary field of view, compact emission dominates the foreground contamination. These two mechanisms imprint a characteristic {\it pitchfork} signature on the ``foreground wedge'' in Fourier delay space. Based on these results, we propose that selective down--weighting of data based on antenna spacing and time can mitigate foreground contamination substantially by a factor $\sim 100$ with negligible loss of sensitivity.

@@

At the end of the recombination epoch, the Universe was completely neutral. This period, referred to as the {\it Dark Ages} in the Universe's history, is characterized by the localized accumulation of matter under the influence of gravity. And it ended with the formation of the first stars and galaxies which started emitting ultra--violet and X--ray radiation, thereby reionizing the neutral medium in their surroundings. This commenced the epoch of reionization (EoR) --- a period of non--linear growth of matter density perturbations and astrophysical evolution. % Studying the EoR holds the key to understanding this evolution.

Observing redshifted 21~cm radiation generated by the spin flip transition of H{\sc i} has been identified as a direct probe of the EoR \citep{sun72,sco90,mad97,toz00,ili02}. Detecting this signal has recently emerged as a very promising experiment to fill the gaps in our understanding of the Universe's history.

@@

Considerable effort is being made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum and formulating robust estimators of the true power spectrum \citep{bow09,liu09,dat10,liu11,mor12,tro12,pob13,thy13,dil14,liu14a,liu14b}. {\bf A model that provides a generic explanation for the observed foreground power spectrum has emerged, whereby the wide--field (and chromatic) response of the instrument causes the power in smooth spectrum foregrounds to occupy higher $k$--modes into the so-called `wedge'}. The conservative foreground strategy, referred to as {\it avoidance}, that has developed alongside this work is to discard $k$--modes which could be contaminated \citep[e.g.,][]{par14}. The more aggressive alternative is to subtract a sky model and regain access to modes that would be discarded by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we focus primarily on extending the {\it avoidance} strategy by identifying foreground components at greatest risk to ``leak'' from foreground modes to EoR modes and proposing a scheme for down--weighting these components.

Foregrounds with intrinsic deviations from spectral smoothness, instruments with high chromaticity, polarization leakage, calibration errors, or approximations in power spectrum analyses can contaminate the true EoR H{\sc i} power spectrum. Here we use existing catalogs and a high fidelity instrumental model to capture both foreground and instrumental chromaticity. To decouple these effects from possible analysis effects, such as those pointed out by \citet{haz13}, we compute power spectra using a {\it \textbf{per-baseline}} approach of \citet{par12b}. This approximates the power spectrum as the inverse Fourier transform of the spectra generated by the instrument's correlator.

{\bf We investigate signatures generic to all wide--field measurements of EoR power spectra and explore antenna shapes that are most desirable from the viewpoint of foreground contamination for future EoR experiments}. After matching observations from the Murchison Widefield Array with instrumental simulations of the entire sky, we examine the {\bf shape of the delay spectrum} of each baseline. We report two important findings: in the {\it avoidance} strategy, the foregrounds that most severely {\bf obscure} the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon; and, diffuse Galactic emission plays a significant role hitherto unpredicted. We quantify these by separating the simulations into components based on type of foreground emission. We then arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to down--weight adversely contaminated baselines.

@@

Interferometer array data known as {\it visibilities}, $V_{bf}\equiv V_{bf}(\boldsymbol{b},f)$, represent correlations between time--series of electric fields measured by different antenna pairs with separation vectors $\boldsymbol{b}$ and then Fourier transformed along the time axis to obtain a spectrum along the frequency ($f$) axis. If $I(\hat{\boldsymbol{s}},f)$ and $A(\hat{\boldsymbol{s}},f)$ are the sky brightness and antenna's directional power pattern, respectively, at different frequencies as a function of direction on the sky denoted by the unit vector ($\hat{\boldsymbol{s}}$), and $W_f\equiv W_f(f)$ denotes instrumental bandpass weights, then $V_{bf}$ can be written as:
\begin{align}\label{eqn:obsvis}
  V_{bf} &= \iint\limits_\textrm{sky} A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)\,W_f\,e^{-i2\pi f\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}}\,\dif\Omega,
\end{align}
where, $c$ is the speed of light, and $\dif\Omega$ is the solid angle element to which $\hat{\boldsymbol{s}}$ is the unit normal vector. This equation is valid {\bf in general, including wide--field measurements and is a slight adaptation from \citet{van34}, \citet{zer38}, and \citet{tho01}}.

@@

Equation~\ref{eqn:obsvis} can be {\bf equivalently expressed} as:
\begin{align}\label{eqn:vis}
  V_{uf} &\equiv V_{uf}(\boldsymbol{u},f) \nonumber\\
  &= \iint\limits_\textrm{sky} A(\hat{\boldsymbol{s}},f)\,I(\hat{\boldsymbol{s}},f)\,W_f\,e^{-i2\pi\boldsymbol{u}\cdot\hat{\boldsymbol{s}}}\,\dif\Omega,
\end{align}
where, $\hat{\boldsymbol{s}}$ is measured with reference to a location on the sky referred to as the {\it phase center}, and $\boldsymbol{u}\equiv (u,v,w)$ denotes the spatial frequency vector. $w$ is aligned parallel to the direction of the phase center, while $u$ and $v$ lie on the transverse plane perpendicular to it. For measurements that are constrained to lie on this plane, we can choose $w=0$ without loss of generality and $\boldsymbol{u}$ effectively reduces to $\boldsymbol{u}\equiv (u,v)$, a two--dimensional vector. Then, $\boldsymbol{u}$ is directly related to the transverse spatial wavenumber mode as: 
\begin{align}\label{eqn:k-perp}
  \boldsymbol{k}_\perp &\equiv \frac{2\pi\boldsymbol{u}}{D}, 
\end{align}
where, $D\equiv D(z)$ is the transverse comoving distance at redshift $z$.

@@

Foregrounds can be described in both the delay spectrum and Fourier frameworks. For our study, we find the former to be simple and yet extremely useful. In order to express a quantity derived from $V_{u\eta}$ that has units identical to that of the cosmological H{\sc i} power spectrum, we define the delay power spectrum:
\begin{align}\label{eqn:fake-power-spectrum}
  P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel) &\equiv |V_{b\tau}|^2\left(\frac{A_\textrm{e}}{\lambda^2\Delta B}\right)\left(\frac{D^2\Delta D}{\Delta B}\right)\left(\frac{\lambda^2}{2k_\textrm{B}}\right)^2,
\end{align}
where, $\boldsymbol{u}$ and $\eta$ in equations~\ref{eqn:k-perp} and \ref{eqn:k-prll} have been correspondingly replaced with $\boldsymbol{b}/\lambda$ and $\tau$ respectively.

@@

{\bf We give a brief overview of some parameters of Fourier space which is generic to all experiments that use a similar approach}. Figure~\ref{fig:fourier-space} illustrates the Fourier space in which the delay (and power) spectra of H{\sc i} from the EoR are presented in this paper. $|\boldsymbol{b}|$ and $k_\perp$, denoting spatial scales in the transverse direction {\bf (tangent plane to the celestial sphere)}, form the $x$--axis. $\tau\approx\eta\propto k_\parallel$, denoting spatial scales along line--of--sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge} \citep{dat10}, whose boundaries are determined by the antenna spacings and the light travel times across them. These boundaries, called horizon delay limits \citep{ved12,par12b}, are shown by solid lines. The spectral transfer function of the instrument convolves the {\it foreground wedge} and stretches it further (unshaded narrow strips bounded by solid and dashed lines) along $\tau$--axis \citep{par12b,thy13}. The width of this narrow strip is inversely proportional to the operating bandwidth. The region of Fourier space excluding the {\it foreground wedge} and the narrow strips is the so--called {\it EoR window}, shown in light and medium shades. {\bf In the context of EoR studies in Fourier space, the H{\sc i} power spectrum from the EoR is expected to decrease rapidly with $|\boldsymbol{k}|$. Hence, the brightest EoR signal will be observed on the shortest baselines and smallest delays. Thus the region of interest for EoR studies relying on {\it avoidance} strategy is just beyond the horizon delay limits (dashed lines) on short baselines, marked as the region of {\it maximal EoR sensitivity}}.

{\bf In the specific case of the MWA, which has a passband constructed using coarse channels, there are period grating responses resulting in repetitions of the {\it foreground wedge} at multiples of 0.78~$\mu$s. Thus with reference to the MWA, the {\it EoR window} lies outside the dashed lines but inside the first grating response ($|\tau| \lesssim 0.78\,\mu$s, dot--dashed lines) and is shown in medium shade}.

@@

We obtain the delay spectrum of visibilities by taking the delay transform of each baseline's spectrum (equation~\ref{eqn:delay-transform}) choosing $W'_f$ to be a {\it Blackman--Harris} window function. The sky spectrum is multiplied by weights, $W_f$, due to the instrumental passband and channels possibly flagged for interference. In delay--space, these weights translate into a convolution by a point spread function (PSF). We deconvolve this PSF using a one dimensional CLEAN algorithm \citep{tay99} as described for the delay axis \citep{par09,par12b} to obtain the final delay spectra. The CLEAN procedure iteratively finds and subtracts peak values convolved by the Fourier transform of the weights. We limit the selection of peaks to modes inside the horizon delay limit, corresponding to smooth spectrum objects in the visible sky hemisphere.

@@

When the synthesized field is small, where $A(\hat{\boldsymbol{s}},f)$ or $I(\hat{\boldsymbol{s}},f)$ is significant only for $|l| \ll 1$ and $|m| \ll 1$, equation~\ref{eqn:obsvis-lmn} reduces to a simple two--dimensional Fourier transform \citep{tay99,tho01} between the apparent sky brightness and measured visibilities. It is in this context that radio interferometers are understood to be sensitive only to fluctuations and not to a uniform sky brightness distribution.

In a wide--field measurement, neither $A(\hat{\boldsymbol{s}},f)$ nor $I(\hat{\boldsymbol{s}},f)$, in general, is negligible anywhere in the visible hemisphere. The solid angle per pixel on the sky in direction cosine coordinates changes significantly with direction (equation~\ref{eqn:solid-angle}), increasing steeply towards the horizon. Hence, the approximations in the narrow--field scenario do not apply. For example, even if $A(\hat{\boldsymbol{s}},f)$ and $I(\hat{\boldsymbol{s}},f)$ are held constant across the visible hemisphere, the amplitude of the integrand in equation~\ref{eqn:obsvis-lmn} is still dependent on direction. Therefore, in a significant departure from a narrow--field measurement, the wide--field visibility from a uniform brightness distribution on a non--zero antenna spacing is not zero.

@@

We interpret this as due to equal--sized delay bins subtending larger solid angles near the horizon thereby containing larger integrated emission. An alternate, but equivalent, interpretation is that baseline vectors (including those with largest lengths) are foreshortened towards the horizon along their orientation. Thus, they become sensitive to larger angular scales that match the inverse of their foreshortened lengths along these directions.

\citet{thy13} found evidence of this feature in their statistical models. In line with our reasoning, they attribute it to a steep rise in solid angles subtended by delay bins near the horizon limits. \citet{pob13} also find a similar ``edge brightening'' feature which they attribute to Galactic plane emission near the horizon. From their discussion, it is unclear what fraction of power in that feature arises from such wide--field effects.

We conclude these are generic to all instruments making wide--field measurements. The nature of the specific instrument used for observing will control the amplitude of these effects, which we explore below.

@@

% \subsubsection{Diffuse Foreground Model}\label{sec:DSM}

For the diffuse component, we use an all--sky radio foreground model \citep{deo08} to estimate the emission at 185~MHz. At this frequency, since this map is predominantly based on the 408~MHz map of \citet{has82} which has an angular resolution of 0.85\arcdeg, we smoothed the 185~MHz map to the same resolution. However, to avoid any artifacts from sampling this map, we sample it at $\approx 27$\arcmin~intervals. We model the diffuse foreground spectra with a unique spectral index at each pixel in the map, estimated from model maps at 170~MHz and 200~MHz.

% \subsubsection{Compact Foreground Model}\label{sec:CSM}

The model described above is primarily a model of the diffuse foreground sky. While it contains faint compact emission blended in with the diffuse emission, {\bf bright point sources have been removed \citep{deo08}}. In order to supplement it with missing bright compact emission, we use classical radio source confusion estimates to determine the nominal flux density threshold and include point sources brighter than this threshold. Slightly different criteria are in common use in radio astronomy to estimate radio source confusion \citep[see Appendix of][and references therein]{thy13}. For an angular resolution of 0.85\arcdeg, using a conservative `$S_\textrm{c}=5\sigma_\textrm{c}$' criterion, we determine the flux density threshold to be $\approx 10$~Jy. Other liberal criteria that yield a lower threshold carry a greater risk of double--counting point sources which might be already blended in with the diffuse sky model.

@@

These criteria yield 100 objects from the SUMSS catalog and 250 objects from the NVSS catalog. Together with the diffuse foreground model, we obtain an all--sky foreground model consisting of both compact and diffuse emission. Figure~\ref{fig:sky-model-generic} shows the diffuse (top) and compact (bottom) foreground emission model used in our study. In this snapshot at 0.09~hours LST, the Galactic center in the diffuse model has just set in the West.

@@

The instrument model consists primarily of the antenna power pattern, $A(\hat{\boldsymbol{s}},f)$ (see equation~\ref{eqn:obsvis}). It is determined by the shape of its aperture. Through the use of a few examples, we examine the role the geometrical shape of the aperture plays in shaping the characteristics of delay power spectrum. We consider the following antenna elements placed at the MWA tile locations:
\begin{itemize}
\item {\it Dipole}: An East--West dipole of length 0.74~m at a height 0.3~m above a ground plane. $A_\textrm{e}=(\lambda/2)^2$.
\item {\it Phased Array}: A 4$\times$4 array of isotropic radiators with a grid spacing of 1.1~m at a height 0.3~m above the ground plane placed in an arrangement similar to that of an MWA tile. $A_\textrm{e}=16\,(\lambda/2)^2$. 
\item {\it Dish}: Diameter of 14~m similar to that proposed for HERA, with $A_\textrm{e}\approx 154$~m$^2$. The power pattern is simulated using an {\it Airy} pattern where its sensitivity beyond the horizon is forced to zero.
\end{itemize}

The power patterns of these antenna geometries at 185~MHz for the {\it zenith} pointing are shown in Figure~\ref{fig:power-patterns}. The presence of a ground plane below the dipole and the phased array introduces a smoother suppression of sensitivity near the horizon, whereas this transition is abrupt in the case of the dish.

@@

The delay power spectra without thermal noise component for these antenna shapes are shown in Figure~\ref{fig:antennas-delay-spectra}. The occupancy of the power patterns on the sky is clearly correlated with that in the delay spectra. Further, the strength of the primary lobe centered on the pointing center is correlated with the delay power spectrum centered on $\tau=0$; and, the overall rate of decrease in the power sensitivity away from the pointing center is correlated with the rate of drop in power away from $\tau=0$.
